---
title: "Snake model residuals"
---

**Purpose:** Explore the relationship between residuals from K. Gauthier's daily stream temperature model and independent metrics of groundwater availability. This is somewhat complicated by the fact that Kevin's model already uses groundwater availability as a covariate, so we really aren't evaluating how much additional information the groundwater metric adds, but rather the variance that isn't explained by the existing metric (e.g., driven by spatial/temporal variation in groundwater temperature, lagged groundwater temperature signals, etc.).


```{r include=FALSE}
library(tidyverse)
library(ggforce)
library(lme4)
library(egg)
library(GGally)
```


## Data

Load data from Kevin's temp model: 
```{r}
dat <- read_csv("Stream temperature/model_obs_pred_data.csv") %>% 
  mutate(model_resid = model_pred - meanT,
         year = as.factor(year(date))) %>%
  group_by(site, year) %>%
  mutate(model_resid_lag = lag(model_resid)) %>%
  ungroup()

head(dat)
length(unique(dat$site))
```


## Summarize error

"Global" RMSE
```{r}
sqrt(mean((dat$model_resid)^2))
```

"Global" mean absolute error
```{r}
mean(abs(dat$model_resid))
```

Plot relationships between model error (RMSE and MAE) and site-specific covariates
```{r fig.width=6, fig.height=12}
errorterms <- dat %>%
  group_by(site) %>%
  summarize(gw_influence = mean(gw_influence),
            h2oAreaKm2_log = mean(h2oAreaKm2_log),
            forest_buff_perc = mean(forest_buff_perc),
            h2o_slope_mean = mean(h2o_slope_mean),
            h2o_elev_mean = mean(h2o_elev_mean),
            sampsize = n(),
            rmse = sqrt(mean((model_resid)^2)),
            mae = mean(abs(model_resid))) %>%
  ungroup()

# RMSE
p1 <- errorterms %>%
  ggplot(aes(x = log(gw_influence), y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("log(Groundwater influence)") + ylab("RMSE") +
  theme_bw() + theme(panel.grid = element_blank())

p2 <- errorterms %>%
  ggplot(aes(x = h2oAreaKm2_log, y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("log(Catchment area)") + ylab("RMSE") +
  theme_bw() + theme(panel.grid = element_blank())

p3 <- errorterms %>%
  ggplot(aes(x = forest_buff_perc, y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Forest cover (% in buffer)") + ylab("RMSE") +
  theme_bw() + theme(panel.grid = element_blank())

p4 <- errorterms %>%
  ggplot(aes(x = h2o_slope_mean, y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Mean catchment slope") + ylab("RMSE") +
  theme_bw() + theme(panel.grid = element_blank())

p5 <- errorterms %>%
  ggplot(aes(x = h2o_elev_mean, y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Mean catchment elevation") + ylab("RMSE") +
  theme_bw() + theme(panel.grid = element_blank())

p6 <- errorterms %>%
  ggplot(aes(x = log(sampsize), y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("log(Sample size)") + ylab("RMSE") +
  theme_bw() + theme(panel.grid = element_blank())


# Mean Absolute Error
p1b <- errorterms %>%
  ggplot(aes(x = log(gw_influence), y = mae)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("log(Groundwater influence)") + ylab("MAE") +
  theme_bw() + theme(panel.grid = element_blank())

p2b <- errorterms %>%
  ggplot(aes(x = h2oAreaKm2_log, y = mae)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("log(Catchment area)") + ylab("MAE") +
  theme_bw() + theme(panel.grid = element_blank())

p3b <- errorterms %>%
  ggplot(aes(x = forest_buff_perc, y = mae)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Forest cover (% in buffer)") + ylab("MAE") +
  theme_bw() + theme(panel.grid = element_blank())

p4b <- errorterms %>%
  ggplot(aes(x = h2o_slope_mean, y = mae)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Mean catchment slope") + ylab("MAE") +
  theme_bw() + theme(panel.grid = element_blank())

p5b <- errorterms %>%
  ggplot(aes(x = h2o_elev_mean, y = mae)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Mean catchment elevation") + ylab("MAE") +
  theme_bw() + theme(panel.grid = element_blank())

p6b <- errorterms %>%
  ggplot(aes(x = log(sampsize), y = mae)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("log(Sample size)") + ylab("MAE") +
  theme_bw() + theme(panel.grid = element_blank())

# combine plots
ggarrange(p1, p1b, 
          p2, p2b, 
          p3, p3b,
          p4, p4b,
          p5, p5b,
          p6, p6b, nrow = 6, ncol = 2, byrow = TRUE)
  
```

View pairs plots among site-specific covariates (some potential multicollinearity)
```{r fig.width=7, fig.height=7}
ggpairs(errorterms %>% select(2:7))
```

Simple linear model quantifying the effect of site-specific covariates on model RMSE
```{r}
summary(lm(rmse ~ gw_influence + h2oAreaKm2_log + forest_buff_perc + h2o_slope_mean + h2o_elev_mean + sampsize, data = errorterms))
```


## Visualize residuals

### Obs. and preds - scatter

```{r}
dat %>%
  ggplot(aes(x = meanT, y = model_pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme_bw() + theme(panel.grid = element_blank()) +
  xlab("Observed temperature") + ylab("Predicted temperature")
```

```{r}
p <- dat %>%
  ggplot(aes(x = meanT, y = model_pred)) +
  geom_point(aes(color = year), alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  theme_bw() + theme(panel.grid = element_blank()) +
  xlab("Observed temperature") + ylab("Predicted temperature") +
  facet_wrap_paginate(~site, ncol = 3, nrow = 4, scales = "free_x")
n_pages(p)
```

::: panel-tabset

#### p1
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 1)
```

#### p2
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 2)
```

#### p3
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 3)
```

#### p4
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 4)
```

#### p5
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 5)
```

#### p6
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 6)
```

#### p7
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 7)
```

:::




### Obs. and preds - time series

View time series with observations (**black text**) and model predictions ([**red text**]{style="color: red;"})
```{r}
p <- dat %>% 
  ggplot() +
  geom_line(aes(x = date, y = meanT)) + 
  geom_line(aes(x = date, y = model_pred), color = "red") +
  theme_bw() +
  facet_wrap_paginate(~site, ncol = 3, nrow = 4, scales = "free_x")
n_pages(p)
```

::: panel-tabset

#### p1
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 1, scales = "free_x")
```

#### p2
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 2, scales = "free_x")
```

#### p3
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 3, scales = "free_x")
```

#### p4
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 4, scales = "free_x")
```

#### p5
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 5, scales = "free_x")
```

#### p6
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 6, scales = "free_x")
```

#### p7
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 7, scales = "free_x")
```

:::


### Residuals - time series

View all time series of residuals together
```{r}
dat %>%
  ggplot(aes(x = DOY, y = model_resid)) + 
  geom_point()
```


Facet by site, color by year:
```{r}
p <- dat %>% 
  ggplot() +
  geom_line(aes(x = DOY, y = model_resid, color = year)) +
  geom_abline(intercept = 0, slope = 0, linetype = "dashed") +
  theme_bw() +
  facet_wrap_paginate(~site, ncol = 3, nrow = 4)
n_pages(p)
```

::: panel-tabset

#### p1
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 1)
```

#### p2
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 2)
```

#### p3
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 3)
```

#### p4
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 4)
```

#### p5
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 5)
```

#### p6
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 6)
```

#### p7
```{r fig.width=8, fig.height=10}
p + facet_wrap_paginate(~ site, ncol = 3, nrow = 4, page = 7)
```

:::


Annual pattern of residuals is generally consistent among years, within sites. Strong autocorrelation



## Model residuals

Absolute residuals by groundwater input
```{r}
summary(lm(abs(model_resid) ~ abs(model_resid_lag) + gw_influence, data = dat))

dat %>%
  ggplot(aes(x = gw_influence, y = abs(model_resid))) +
  geom_point() +
  theme(legend.position = "none") +
  geom_smooth(method = "lm")
```


Residuals by groundwater and month
```{r fig.width=7, fig.height=7}
dat %>%
  ggplot(aes(x = gw_influence, y = model_resid)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~tim.month) + 
  theme_bw()
```


Explore time varying effects of groundwater on model residuals (linear mixed effects model includes lagged/autocorrelation term and random intercepts for year).
```{r}
# fit mixed effects models in a loop, independently for each month: residuals as a function of lagged residuals, site-level groundwater influence, and random intercepts for year
modlist <- list()
for (i in 1:12) {
  modlist[[i]] <- lmer(model_resid ~ model_resid_lag + gw_influence + (1|year), data = dat %>% filter(tim.month == i))
}

# empty tibble to store coefficients
efftib <- tibble(month = 1:12, 
                 mean.gw = rep(NA, 12),
                 lower.gw = rep(NA, 12),
                 upper.gw = rep(NA, 12), 
                 
                 mean.ac = rep(NA, 12),
                 lower.ac = rep(NA, 12),
                 upper.ac = rep(NA, 12))

# get coefficients, means and 95% confidence intervals
for (i in 1:12) {
  efftib$mean.gw[i] <- fixef(modlist[[i]])[3]
  efftib$lower.gw[i] <- confint(modlist[[i]])[5,1]
  efftib$upper.gw[i] <- confint(modlist[[i]])[5,2]
  
  efftib$mean.ac[i] <- fixef(modlist[[i]])[2]
  efftib$lower.ac[i] <- confint(modlist[[i]])[4,1]
  efftib$upper.ac[i] <- confint(modlist[[i]])[4,2]
}

# plot autocorrelation terms
efftib %>%
  ggplot() +
  geom_point(aes(x = month, y = mean.ac)) +
  geom_errorbar(aes(x = month, ymin = lower.ac, ymax = upper.ac), width = 0.4) +
  geom_abline(intercept = 0, slope = 0, linetype = "dashed") +
  theme_bw() + theme(panel.grid = element_blank()) + 
  ylab("Autocorrelation term") +
  scale_x_continuous(breaks = c(1:12))

# plot groundwater effects
efftib %>%
  ggplot() +
  geom_point(aes(x = month, y = mean.gw)) +
  geom_errorbar(aes(x = month, ymin = lower.gw, ymax = upper.gw), width = 0.4) +
  geom_abline(intercept = 0, slope = 0, linetype = "dashed") +
  theme_bw() + theme(panel.grid = element_blank()) + 
  ylab("Effect of groundwater on model residuals") +
  scale_x_continuous(breaks = c(1:12))
  
```


## Model RMSE

Alternatively, we can model monthly RMSE, rather than the raw residuals. This eliminates the need for an autocorrelation term and is based on a coarser measure of predictive accuracy (is using/modeling all residuals pseudoreplication), but limits our ability to discriminate between under vs over-estimates/predictions. 

### Monthly models

Summarize data by site, year, and month
```{r}
errorterms <- dat %>%
  group_by(site, year, tim.month) %>%
  summarize(gw_influence = mean(gw_influence),
            h2oAreaKm2_log = mean(h2oAreaKm2_log),
            forest_buff_perc = mean(forest_buff_perc),
            h2o_slope_mean = mean(h2o_slope_mean),
            h2o_elev_mean = mean(h2o_elev_mean),
            sampsize = n(),
            rmse = sqrt(mean((model_resid)^2)),
            mae = mean(abs(model_resid))) %>%
  ungroup() %>%
  filter(sampsize >= 28)
```

RMSE by groundwater and month
```{r fig.width=7, fig.height=7}
errorterms %>%
  ggplot(aes(x = gw_influence, y = rmse)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~tim.month) + 
  theme_bw()
```

Fit linear mixed effects models in a loop, independently for each month: RMSE as a function of site-level groundwater influence and random intercepts for year
```{r}
modlist <- list()
for (i in 1:12) {
  modlist[[i]] <- lmer(rmse ~ gw_influence + (1|year), data = errorterms %>% filter(tim.month == i))
}

# empty tibble to store coefficients
efftib <- tibble(month = 1:12, 
                 mean.gw = rep(NA, 12),
                 lower.gw = rep(NA, 12),
                 upper.gw = rep(NA, 12))

# get coefficients, means and 95% confidence intervals
for (i in 1:12) {
  efftib$mean.gw[i] <- fixef(modlist[[i]])[2]
  efftib$lower.gw[i] <- confint(modlist[[i]])[4,1]
  efftib$upper.gw[i] <- confint(modlist[[i]])[4,2]
}

# plot groundwater effects
efftib %>%
  ggplot() +
  geom_point(aes(x = month, y = mean.gw)) +
  geom_errorbar(aes(x = month, ymin = lower.gw, ymax = upper.gw), width = 0.4) +
  geom_abline(intercept = 0, slope = 0, linetype = "dashed") +
  theme_bw() + theme(panel.grid = element_blank()) + 
  ylab("Effect of groundwater on model RMSE") +
  scale_x_continuous(breaks = c(1:12))
```


### Interaction model

Same, but different: model the interaction between month and groundwater directly:
```{r}
# errorterms %>% 
#   group_by(tim.month) %>%
#   summarize(rmse = mean(rmse)) %>%
#   ggplot(aes(x = tim.month, y = rmse)) +
#   geom_point() +
#   geom_line() +
#   theme_bw() + theme(panel.grid = element_blank()) +
#   xlab("Month") + ylab("Mean RMSE") +
#   scale_x_continuous(breaks = c(1:12))

errorterms %>% 
  #filter(gw_influence < 0.15) %>%
  ggplot(aes(x = tim.month, y = rmse, color = gw_influence)) + 
  geom_point() +
  geom_line(aes(group = interaction(site, year), color = gw_influence)) +
  theme_bw() + theme(panel.grid = element_blank()) +
  xlab("Month") + ylab("RMSE") +
  scale_x_continuous(breaks = c(1:12))
```

Based on the plot of the data above, fit a linear interaction model with second order polynomial for groundwater influence:
```{r}
mymod <- lm(rmse ~ gw_influence + tim.month + I(tim.month^2) + gw_influence*tim.month + gw_influence*I(tim.month^2), data = errorterms)
summary(mymod)
```

Plot the model predictions:
```{r}
mymonths <- seq(from = 1, to = 12, length.out = 100)
preds_gw0 <- predict(mymod, newdata = list(gw_influence = rep(0, times = 100), tim.month = mymonths),
                     interval = "confidence", level = 0.95)
preds_gw1 <- predict(mymod, newdata = list(gw_influence = rep(1, times = 100), tim.month = mymonths),
                     interval = "confidence", level = 0.95)

ggplot() + 
  geom_ribbon(aes(x = mymonths, ymin = preds_gw0[,2], ymax = preds_gw0[,3]), 
              color = NA, fill = "grey30", alpha = 0.3) +
  geom_ribbon(aes(x = mymonths, ymin = preds_gw1[,2], ymax = preds_gw1[,3]), 
              linewidth = 1, color = NA, fill = "dodgerblue3", alpha = 0.3) +
  geom_line(aes(x = mymonths, y = preds_gw0[,1]), linewidth = 1, color = "grey30") +
  geom_line(aes(x = mymonths, y = preds_gw1[,1]), linewidth = 1, color = "dodgerblue3") +
  theme_bw() + theme(panel.grid = element_blank()) +
  xlab("Month") + ylab("RMSE") +
  scale_x_continuous(breaks = c(1:12))
```


Derive the month-specific effect of groundwater on RMSE
```{r}
mygw <- seq(from = 0, to = 1, length.out = 100)
predlist <- list()
for (i in 1:12) {
  preds <- predict(mymod, newdata = list(gw_influence = mygw, tim.month = rep(i, times = 100)),
                   interval = "confidence", level = 0.95)
  predlist[[i]] <- tibble(month = rep(i, times = 100),
                          gw = mygw,
                          fit = preds[,1],
                          lwr = preds[,2],
                          upr = preds[,3])
}
predtib <- bind_rows(predlist)

predtib %>%
  ggplot(aes(x = gw, y = fit, group = as.factor(month), color = as.factor(month))) +
  geom_line(linewidth = 1) +
  theme_bw() + theme(panel.grid = element_blank()) +
  xlab("Groundwater influence") + ylab("RMSE")
```


*Note that these results generally match those above (results from the monthly RMSE models), but packages everything into a single model statement.*
